---
title: "SIR models with odin, dust and mcstate"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{sir_models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`mcstate` sits at the top of a stack of packages, which together allow inference of a stochastic model's parameters by fitting to a time-series of observational data. To use `mcstate`, we require the following pieces:

- A generative model for a state-space process, implemented in [`dust`](https://mrc-ide.github.io/dust/). This will generate (stochastic) updates of the state at each time step.
- A time-series of observed data, which may include from multiple sources (e.g. cases, hospital beds, deaths).
- A comparison function, which calculates the likelihood of a model state, given the observed data.
- Optionally an index function, which identifies the indices of the state used to compute the comparison. This is particularly useful when the state space is large, and only a small number of states are used in the comparison function.

Below we describe a typical use case for infectious disease: given daily case counts during an epidemic, how can we estimate key features of the transmission process such as the transmission rate, recovery rate and from these the basic reproductive number $R_0$.

# Stochastic SIR model definition

A simple definition of the SIR model, as given in the [odin documentation](https://mrc-ide.github.io/odin/articles/discrete.html) is:
$$\begin{align*} 
\frac{dS}{dt} &= -\beta \frac{SI}{N} \\
\frac{dI}{dt} &= \beta \frac{SI}{N} - \gamma I \\
\frac{dR}{dt} &= \gamma I \\
\end{align*}$$
$S$ is the number of susceptibles, $I$ is the number of infected and $R$ is the number recovered; the total population size $N = S + I + R$ is constant. $\beta$ is the infection rate, $\gamma$ is the recovery rate.

Discretising this model in time steps of width $dt$ gives the following update equations for each time step:

$$\begin{align*} 
S_{t+1} &= S_t - n_{SI} \\
I_{t+1} &= I_t + n_{SI} - n_{IR} \\
R_{t+1} &= R_t - n_{IR}
\end{align*}$$

where
$$\begin{align*} 
n_{SI} &\sim B(S, 1 - e^{-\beta \frac{I}{N} \cdot dt)} \\
n_{IR} &\sim B(I, 1 - e^{-\gamma \cdot dt)}
\end{align*}$$

This example is included with the `dust` package, and can be loaded with:

```{r}
gen_sir <- dust::dust_example("sir")
```

To learn how to implement this yourself, see the vignette over at [`odin.dust`](https://mrc-ide.github.io/odin.dust/).

# Inferring parameters with mcstate

Typically these models will be used in conjunction with observed data to infer the highest posterior density values for model parameters, most likely trajectories up to the present (nowcast), and make forecasts using these parameters. The `mcstate` package implements particle filter and [pMCMC](https://dx.doi.org/10.1016/j.epidem.2019.100363) methods to make all of this this possible from `dust` models.

A comparison function is needed to define a likelihood, i.e. the probability of observing this trajectory, given the data. This is combined with a prior to generate a posterior distribution for each parameter, which in the SIR model gives the distribution of likely values for $\beta$ and $\gamma$. 

This is straightforward for a deterministic model, but for a stochastic model over a long time-series, trajectories can quickly become inconsistent with the data, and the massive space of potential outcomes makes downstream sampling highly inefficient. `mcstate` therefore implements a particle filter, also known as sequential Monte Carlo (SMC), which runs multiple trajectories, and at each stage where data is available resamples them in proportion to their likelihood, keeping only those trajectories close to the data, and with likelihoods that are plausibly important parts of the posterior. Importantly, this technique produces an unbiased estimate of the likelihood for a given set of parameters.

The anatomy of an mcstate particle filter, as noted above, consists of three main components:

- A set of observations to fit the model to, generated using `mcstate::particle_filter_data()`.
- A model to fit, which must be a dust generator, either `dust::dust()` or `odin.dust::odin_dust()`.
- A comparison function, which is an R function which calculates the likelihood of the state given the data at one time point.

## Model data

To demonstrate this, we will attempt to infer the parameters $\beta$ and $\gamma$ using daily case counts. In reality these counts would be measured, but here we generate them from a single run of the model, and adding some constant noise proportional to the epidemic size:

```{r}
# Set up model
dt <- 0.25
inv_dt <- 1/dt
sir <- gen_sir$new(data = list(dt = dt, S_ini = 1000, I_ini = 10, beta = 0.2, gamma = 0.1),
                   step = 0,
                   n_particles = 1L,
                   n_threads = 1L,
                   seed = 2L)

# Run model forward 100 days, using 4 steps a day
# Calculate the number of new cases by counting how many susceptibles have been depleted between updates
n_steps <- 100
incidence <- rep(NA, length.out = n_steps)
true_history <- array(NA_real_, c(sir$info()$len, 1, n_steps + 1))
true_history[, 1, 1] <- sir$state()
for (t in seq_len(n_steps)) {
    state_start <- sir$state()
    state_end <- sir$run(t * inv_dt)
    true_history[, 1, t + 1] <- state_end
    incidence[t] <- state_start[2, 1] - state_end[2, 1]
}

# Add very simple form of noise
incidence <- incidence + 
  round(rnorm(mean = 0, n = length(true_history[3,1,-1]), sd = 0.1*sqrt(true_history[3,1,-1])))
incidence <- unlist(lapply(incidence, max, 0))
incidence_data <- data.frame(cases = incidence, day = seq_len(n_steps))
incidence
```

This can then be processed with `particle_filter_data()` as follows:

```{r paged.print=TRUE, layout="l-body-outset"}
sir_data <- particle_filter_data(data = incidence_data,
                                 time = "day", 
                                 rate = inv_dt)
paged_table(sir_data)
```

More complex models may have additional data streams added as further columns (e.g. deaths, hospital admissions). This will be fed as observed data to the compare function, which will combine and compute the evidence for the current model state.

## Defining the comparison function

The comparison function will typically be a likelihood, the probability of the simulation run given the data. This can be any function in R which takes the current simulated `state`, the state at the previous time step `prev_state`, a `data` row for the current time point and a list of any parameters `pars` the function needs.

For this model the number of cases can be calculated from the change in susceptibles between time steps (as above). In this comparison function, the number of cases within each time interval is modelled as being Poisson distributed. A small amount of noise is added to prevent zero expectations.

```{r}
case_compare <- function(state, prev_state, observed, pars = NULL) {
  if (is.na(observed$cases)) {
    return(NULL)
  }
  if (is.null(pars$compare$exp_noise)) {
    exp_noise <- 1e6
  } else {
    exp_noise <- pars$compare$exp_noise
  }
  
  incidence_modelled <-
     prev_state[2, , drop = TRUE] - state[2, , drop = TRUE]
  incidence_observed <- observed$cases
  lambda <- incidence_modelled +
    rexp(n = length(incidence_modelled), rate = exp_noise)
  dpois(x = incidence_observed, lambda = lambda, log = TRUE)
}
```

## Inferring parameters

Using these pieces, we can set up a particle filter as follows:

```{r}
n_particles <- 100
filter <- particle_filter$new(data = sir_data,
                              model = gen_sir,
                              n_particles = n_particles,
                              compare = case_compare,
                              n_threads = 4L,
                              seed = 1L)
```

We can now run the particle filter forward, which will run and resample 100 trajectories, and return the final likelihood. It is important to set the correct time-step $dt$ here, as we are using 0.25, rather than the default value of 1 defined in the odin model.

```{r}
filter$run(save_history = TRUE, pars = list(dt = dt))
```

If we plot these along with the data, we can see compared to above that only trajectories consistent with the data are kept, and the variance between particles has been reduced:

```{r fig.height=5, fig.width=7}
history <- filter$history()

par(mar = c(4.1, 5.1, 0.5, 0.5), las = 1)
matplot(incidence_data$day, t(history[2, , -1]), type = "l",
        xlab = "Time", ylab = "Number of individuals",
        col = "#999966", lty = 1, ylim = range(history))
matlines(incidence_data$day, t(history[3, , -1]), col = "#8c8cd9", lty = 1)
matlines(incidence_data$day, t(history[4, , -1]), col = "#cc0044", lty = 1)
matpoints(incidence_data$day, t(true_history[2:4, , -1]), pch = 19,
          col = c("#999966", "#8c8cd9", "#cc0044"))
legend("left", lwd = 1, col = c("#999966", "#8c8cd9", "#cc0044"), 
       legend = c("S", "I", "R"), bty = "n")
```

Doing the same with a $\beta$ and $\gamma$ which are further from the truth yields a lower likelihood:

```{r}
filter$run(save_history = TRUE, pars = list(dt = dt, beta = 0.4, gamma = 0.2))
```


```{r fig.height=5, fig.width=7}
history <- filter$history()
matplot(incidence_data$day, t(history[2, , -1]), type = "l",
        xlab = "Time", ylab = "Number of individuals",
        col = "#999966", lty = 1, ylim = range(history))
matlines(incidence_data$day, t(history[3, , -1]), col = "#8c8cd9", lty = 1)
matlines(incidence_data$day, t(history[4, , -1]), col = "#cc0044", lty = 1)
matpoints(incidence_data$day, t(true_history[2:4, , -1]), pch = 19,
          col = c("#999966", "#8c8cd9", "#cc0044"))
legend("left", lwd = 1, col = c("#999966", "#8c8cd9", "#cc0044"), 
       legend = c("S", "I", "R"), bty = "n")
```

Typically you will not use the `run()` function of the particle filter directly, and instead use the functions below to infer parameters.

### Using MCMC to infer parameters

Using this, we can do a MCMC using the Metropolis-Hastings algorithm to infer the values of $\beta$ and $\gamma$ from daily case counts. Let's first describe the parameters we wish to infer, giving a minimum, maximum. For $\gamma$ we will also apply a gamma prior with shape 1 and rate 0.2: $\Gamma(1, 0.2)$:

```{r}
beta <- pmcmc_parameter("beta", 0.2, min = 0)
gamma <- pmcmc_parameter("gamma", 0.1, min = 0, 
                         prior = function(p) {
                           dgamma(p, shape = 1, scale = 0.2, log = TRUE)
                         })

proposal_matrix <- matrix(c(0.01, 0, 0, 0.01), nrow = 2, ncol = 2, byrow = TRUE)
mcmc_pars <- pmcmc_parameters$new(list(beta = beta, gamma = gamma), proposal_matrix)
```

The proposals for each MCMC step are also set here. Proposals are drawn from a multi-variate normal distribution, with the variance-covariance matrix set here. In this case we have $\beta_{n+1} \sim N(\beta_{n}, 0.01^2)$ and $\gamma_{n+1} \sim N(\gamma_{n}, 0.01^2)$ i.e. no covariance.

Finally, the sampler can be run using these parameters, and the particle filter defined above. Taking 2000 steps with two independent chains:

```{r pmcmc_combined}
pmcmc_run <- 
  pmcmc(
    mcmc_pars,
    filter,
    2000,
    save_state = TRUE,
    save_trajectories = TRUE,
    progress = TRUE,
    n_chains = 2
  )
```

We also provide some basic tools for dealing with this output, before running prediction (see below). We can remove burn-in and thin the four chains as follows:

```{r}
processed_chains <- pmcmc_thin(pmcmc_run, burnin = 500, thin = 2)
processed_chains
apply(processed_chains$pars, 2, mean)
```

### Optimal use of parallelisation

Note that we run four separate chains in serial here. If you have $m$ nodes available, each with $n$ cores, the most efficient way to run this step is to set `nthreads = n` on the filters, and run $m$ independent chains, one on each node setting `n_chains = 1`.

For threads to take effect, `openmp` must be enabled in the compilation step, which you can check by running:

```{r}
gen_sir$public_methods$has_openmp()
```

These independent results can then be collected using `pmcmc_combine()`:

```{r pmcmc_serial}
pmcmc1 <- 
  pmcmc(
    mcmc_pars,
    filter,
    2000,
    save_state = TRUE,
    save_trajectories = FALSE,
    progress = TRUE,
    n_chains = 1
  )
pmcmc2 <- 
  pmcmc(
    mcmc_pars,
    filter,
    2000,
    save_state = TRUE,
    save_trajectories = FALSE,
    progress = TRUE,
    n_chains = 1
  )
pmcmc3 <- 
  pmcmc(
    mcmc_pars,
    filter,
    2000,
    save_state = TRUE,
    save_trajectories = FALSE,
    progress = TRUE,
    n_chains = 1
  )
pmcmc_distributed_run <- pmcmc_combine(pmcmc1, pmcmc2, pmcmc3)
```

The resulting objects can also be analysed in a typical MCMC analysis package such as [`coda`](https://cran.r-project.org/web/packages/coda/index.html). 

```{r coda_plots, fig.height=10, fig.width=14}
library(coda)
mcmc1 <- as.mcmc(cbind(pmcmc1$probabilities, pmcmc1$pars))
mcmc2 <- as.mcmc(cbind(pmcmc2$probabilities, pmcmc2$pars))
mcmc3 <- as.mcmc(cbind(pmcmc3$probabilities, pmcmc3$pars))
mcmc_sir <- mcmc.list(mcmc1, mcmc2, mcmc3)

# The Rhat statistic may not always be possible to compute
# gelman.diag(mcmc_sir)
rejectionRate(mcmc_sir)
effectiveSize(mcmc_sir)

# Remove burn in and thin
mcmc_sampled <- lapply(list(pmcmc1, pmcmc2, pmcmc3), pmcmc_thin, burnin = 1000, thin = 2)
mcmc1 <- as.mcmc(cbind(mcmc_sampled[[1]]$probabilities, mcmc_sampled[[1]]$pars))
mcmc2 <- as.mcmc(cbind(mcmc_sampled[[2]]$probabilities, mcmc_sampled[[2]]$pars))
mcmc3 <- as.mcmc(cbind(mcmc_sampled[[3]]$probabilities, mcmc_sampled[[3]]$pars))
mcmc_sir <- mcmc.list(mcmc1, mcmc2, mcmc3)

summary(mcmc_sir)
plot(mcmc_sir)
```


## Fitting to multiple datastreams

The set up of the particle filter and observation function makes it easy to fit to multiple datastreams within a Bayesian framework. Let us model a disease where a proportion of infected cases $p_{\mathrm{death}} = 0.05$ die rather than recover by adding the following lines to the odin code:

```r
update(S) <- S - n_SI
update(I) <- I + n_SI - n_IR - n_ID
update(R) <- R + n_IR
update(D) <- D + n_ID

n_IR <- rbinom(I, p_IR * dt * (1 - p_death))
n_ID <- rbinom(I, p_IR * dt * p_death)
n_SI <- rbinom(S, p_SI * dt)

initial(D) <- 0
p_death <- user(0.05)
```

```{r}
gen_death <- odin_dust("deaths.R")
```

We can simulate some data to fit to, as before. We will also make the case data patchy, making it missing on some days:

```{r layout="l-body-outset", paged.print=TRUE }
# Set up model
dt <- 0.25
inv_dt <- 1/dt
sir_death <- gen_death$new(data = list(dt = dt, S_ini = 1000, I_ini = 10, beta = 0.2, gamma = 0.1),
                   step = 0,
                   n_particles = 1L,
                   n_threads = 1L,
                   seed = 2L)

# Run model forward 100 days, using 4 steps a day
# Calculate the number of new cases by counting how many susceptibles have been depleted between updates
n_steps <- 100
incidence <- rep(NA, length.out = n_steps)
deaths <- rep(NA, length.out = n_steps)
true_history_deaths <- array(NA_real_, c(sir_death$info()$len, 1, n_steps + 1))
true_history_deaths[, 1, 1] <- sir_death$state()
for (t in seq_len(n_steps)) {
    state_start <- sir_death$state()
    state_end <- sir_death$run(t * inv_dt)
    true_history_deaths[, 1, t + 1] <- state_end
    incidence[t] <- state_start[2, 1] - state_end[2, 1]
    deaths[t] <- state_end[5, 1] - state_start[5, 1]
}

# Add very simple form of noise
missing <- rbinom(n = n_steps, prob = 0.05, size = 1) == 1
incidence[missing] <- NA
combined_data <- particle_filter_data(
  data.frame(cases = incidence, deaths = deaths, day = seq_len(n_steps)),
  time = "day",
  rate = inv_dt
)
paged_table(combined_data)
```

We then need to add this into our comparison function, and in doing so we will also demonstrate the use of the `index` parameter for the particle filter. The only model outputs needed from the comparison function are $S$ and $D$. For models with a very large state space returning only those compartments required by the comparison function can be important for efficiency. An index function can be defined, which operates on dust generators `info()`:

```{r}
index <- function(info) {
  list(run = c(info$index$S, info$index$D), state = c(info$index$S, info$index$I, info$index$R, info$index$D))
}

index(sir_death$info())
```

Which sends the comparison function the indices listed under `run`, in this case the number of susceptibles, and saves any compartments listed under `state` if the history is kept. In the `compare()` function the index of `state` now corresponds to `index$run`. The missing data must be checked for, and contribute 0 to the likelihood if found:

```{r}
combined_compare <- function(state, prev_state, observed, pars = NULL) {
  if (is.na(observed$cases) && is.na(observed$deaths)) {
    return(NULL)
  }
  if (is.null(pars$compare$exp_noise)) {
    exp_noise <- 1e6
  } else {
    exp_noise <- pars$compare$exp_noise
  }
  
  ll_pois <- function(obs, model, exp_noise) {
    if (is.na(obs)) {
      ll_obs <- numeric(length(model))
    } else {
      lambda <- model +
        rexp(n = length(model), rate = exp_noise)
      ll_obs <- dpois(x = obs, lambda = lambda, log = TRUE)
    }
    ll_obs
  }
  
  ll_cases <- ll_pois(observed$cases, 
                      prev_state[1, , drop = TRUE] - state[1, , drop = TRUE],
                      exp_noise)
  ll_deaths <- ll_pois(observed$deaths, 
                      state[2, , drop = TRUE] - prev_state[2, , drop = TRUE],
                      exp_noise)
  
  ll_cases + ll_deaths
}
```

This can be used to set up a particle filter as before:

```{r}
n_particles <- 100
filter <- particle_filter$new(data = combined_data,
                              model = gen_death,
                              n_particles = n_particles,
                              compare = combined_compare,
                              index = index,
                              n_threads = 4L,
                              seed = 1L)
filter$run(save_history = FALSE, pars = list(dt = dt))
```

and this can be used to run the `pmcmc()` method, as above.

## Running predictions

Going back to the simple SIR model above, we can easily continue running the fitted model forward in time using the posterior estimated by `pmcmc()` to create a forecast past the end of the data:

```{r fig.height=5, fig.width=7}
forecast <- predict(processed_chains, 
                    steps = seq(400, 800, 4), 
                    n_threads = 4L, 
                    prepend_trajectories = TRUE,
                    seed = processed_chains$predict$seed)

# Sample 10 particles
mini_forecast <- forecast$state[,sample.int(dim(forecast$state)[2], size = 10),]
time <- mini_forecast[1,1,-1]

matplot(time*dt, t(mini_forecast[2, , -1]), type = "l",
        xlab = "Time", ylab = "Number of individuals",
        col = "#999966", lty = 1, ylim = range(mini_forecast))
matlines(time*dt, t(mini_forecast[3, , -1]), col = "#8c8cd9", lty = 1)
matlines(time*dt, t(mini_forecast[4, , -1]), col = "#cc0044", lty = 1)
matpoints(incidence_data$day, t(true_history[2:4, , -1]), pch = 19,
          col = c("#999966", "#8c8cd9", "#cc0044"))
legend("left", lwd = 1, col = c("#999966", "#8c8cd9", "#cc0044"), 
       legend = c("S", "I", "R"), bty = "n")
```

Up to $t = 100$ is the nowcast, from $t = 100$ to $t = 200$ is the forecast.
